Typed Lua: An Optional Type System for Lua

### 2014-03-12 doc/paper/typedlua.tex
reorganization breaking the subsections of section 2 into individual sections; rewrite of section on modules

Lua is a small imperative scripting language that prefers to provide
mechanisms instead of fixed policies for structuring programs,
due to its history as an embedded extension language. Even features
such as a module system and object orientation are a matter of 
convention instead of built-in features.

The lack of standard policies
is a challenge for the design of a static type system for the Lua
language, and our approach was to extract the most common
idioms from a large corpus of Lua libraries as a guide to what
Typed Lua should focus on, and setting fixed policies for organizing
a program in modules and using class-based object orientation.

Typed Lua allows statically typed Lua code to coexist and interact with 
dynamically typed code. The Typed Lua compiler warns the programmer
about type errors, but always generates Lua code that runs in unmodified
Lua implementations. The programmer can enjoy some of the benefits
of static types even without converting existing Lua modules to typed Lua:
a dynamically typed module can export a statically typed interface, and
statically typed users of the module will have their use of the module
checked by the compiler.

@@

In Section~\ref{sec:proposal} we outline the key parts in
the design of Typed Lua. In Section~\ref{sec:review} we review
related work on mixing static and dynamic typing in the same language.
Finally, Section~\ref{sec:con} gives some concluding remarks and
our plans for the future of Typed Lua.

\section{Typed Lua} \label{sec:proposal}

Lua is an imperative scripting language with first-class functions
(with proper lexical scoping) where the main data structure is the
{\em table}, an associative array that can play the part of arrays,
records, maps, objects, etc., with syntactic sugar and
metaprogramming through operator overloading built into
the language. Unlike other scripting languages, Lua has very
limited coercion among different data types.

The primary use of Lua has always been as an embedded language
for configuration and extension of other applications, and the result
is a fragmented ecosystem of libraries, and different ideas among
Lua programmers on how they should use the language features and
how they should structure programs.

The design of Typed Lua is informed by a (mostly automated) survey
of Lua idioms used in a large corpus of Lua libraries, instead of
relying just on the semantics of the language. We cover the main parts
of the design, along with how they relate to Lua, in the rest of this section,
and discuss some of what was left out in the current version of the design
in the conclusion of the paper.

\subsection{Atomic Types and Functions}

@@

\subsection{Unions}

@@

\subsection{Tables and Interfaces}

@@

\subsection{Modules}

Typed Lua will also introduce a specific syntax for defining modules.
For now, each file is a module.
We use all global names and their types to build the interface that
represents the type of the module, in case the module is not returned
at the end of the file.

\subsection{Objects and Classes}

>>>>

Lua is a small imperative scripting language with first-class
functions (with proper lexical scoping) where the main data
structure is the {\em table}, an associative array that can
play the part of arrays, records, maps, objects, etc.
with syntactic sugar and metaprogramming through operator overloading built into
the language. Unlike other scripting languages, Lua has very
limited coercion among different data types.

The primary use of Lua has always been as an embedded language
for configuration and extension of other applications.
Lua prefers to provide mechanisms instead of fixed policies
for structuring programs, and even features
such as a module system and object orientation are a matter of 
convention instead of built into the language.
The result is a fragmented ecosystem of libraries, and
different ideas among Lua programmers on how they should use
the language features and how they should structure programs.

The lack of standard policies
is a challenge for the design of a static type system for the Lua
language. The design of Typed Lua is informed by a (mostly
automated) survey of Lua idioms used in a large corpus of Lua
libraries, instead of relying just on the semantics of the
language.

Typed Lua allows statically typed Lua code to coexist and
interact with  ynamically typed code. The Typed Lua compiler
warns the programmer about type errors, but always generates
Lua code that runs in unmodified Lua implementations. The
programmer can enjoy some of the benefits of static types even
without converting existing Lua modules to typed Lua:
a dynamically typed module can export a statically typed
interface, and statically typed users of the module will have
their use of the module checked by the compiler.

@@

We cover the main parts of the design, along with
how they relate to Lua, in Sections~\ref{sec:atomic}
through~\ref{sec:classes}. Section~\ref{sec:review} reviews
related work on mixing static and dynamic typing in the same
language. Finally, Section~\ref{sec:con} gives some
concluding remarks and future extensions for Typed Lua.

\section{Atomic Types and Functions}
\label{sec:atomic}

@@

\section{Unions}
\label{sec:unions}

@@

\section{Tables and Interfaces}
\label{sec:tables}

@@

\section{Modules}
\label{sec:modules}

Lua's module system, like other parts of the language, is
a matter of convention. When Lua first needs to load a
module, it executes the module's source file as a function;
the value that this function returns is the module, and Lua
caches it for future loads. While a module can be any
Lua value, most modules are tables where the fields
of the table are functions and other values that the
module exports.

The modules that we surveyed build this table using
three distinct styles. In the first style, the module's
source file ends
with a {\tt return} statement that uses a table constructor
with the exported members. In the second style, the
module declares an empty table in the beginning of its
source file, adds exported members to this table
throughout the module, and returns this table at the end.

These two styles are straightforward for Typed Lua, which
can just take the type of the first value that the module
returns and use it as the type of the module.

In the third style, which has been deprecated in the current
version of Lua, a module begins with a call to the {\tt module}
function. This function installs a fresh table as the
global environment for the rest of the module, so any
assignments to global variables are field assignments to
this table. The {\tt module} function also sets an {\tt \_M}
field in this table as a circular reference to the table
itself, so the module can end with {\tt return \_M}, but
this explicit return is not necessary.

While this style has been deprecated, our survey indicated
that around a third of Lua modules in a popular module
repository still use this style, so Typed Lua also
supports this style: it treats accesses to global
variables as field accesses to an open table
in the top-level scope.

\section{Objects and Classes}
\label{sec:classes}
